---
title: "Filter observations"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    ace_theme: nord_dark
runtime: shiny_prerendered
description: >
  Learn how to filter observations in a data frame. Use `filter()` to extract
  observations from a data frame, and use `&`, `|`, and `!` to write logical
  tests.
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)

tutorial_options(
  exercise.timelimit = 60,
  # A simple checker function that just returns the message in the check chunk
  exercise.checker = function(check_code, ...) {
    list(
      message = eval(parse(text = check_code)),
      correct = logical(0),
      type = "info",
      location = "append"
    )
  }
)
knitr::opts_chunk$set(error = TRUE)
```

## Welcome

This is a demo tutorial. Compare it to the [source code](https://github.com/rstudio/learnr/tree/main/inst/tutorials/ex-data-filter/ex-data-filter.Rmd) that made it.

###

In this tutorial, you will learn how to:

* use `filter()` to extract observations from a data frame or tibble
* write logical tests in R
* combine logical tests with Boolean operators
* handle missing values within logical tests

The readings in this tutorial follow [_R for Data Science_](http://r4ds.had.co.nz/), section 5.2.

### Prerequisites

To practice these skills, we will use the `flights` data set from the nycflights13 package. This data frame comes from the US [Bureau of Transportation Statistics](http://www.transtats.bts.gov/DatabaseInfo.asp?DB_ID=120&Link=0) and contains all `r format(nrow(nycflights13::flights), big.mark = ",")` flights that departed from New York City in 2013. It is documented in `?flights`.

We will also use the ggplot2 package to visualize the data.

If you are ready to begin, click on!

## Filter rows with `filter()`

### filter()

`filter()` lets you use a logical test to extract specific rows from a data frame. To use `filter()`, pass it the data frame followed by one or more logical tests. `filter()` will return every row that passes each logical test.

So for example, we can use `filter()` to select every flight in flights that departed on January 1st. Click Run Code to give it a try:

```{r filter1, exercise = TRUE, exercise.eval = FALSE}
filter(mtcars, carb == 2)
```


###  output

Like all dplyr functions, `filter()` returns a new data frame for you to save or use. It doesn't overwrite the old data frame.

If you want to save the output of `filter()`, you'll need to use the assignment operator, `<-`.

Rerun the command in the code chunk below, but first arrange to save the output to an object named `jan1`.

```{r filter2, exercise = TRUE, exercise.eval = FALSE}
filter(mtcars, carb == 2)
```

```{r filter2-solution}
jan1 <- filter(mtcars, carb == 2)
```

###

Good job! You can now see the results by running the name jan1 by itself. Or you can pass `jan1` to a function that takes data frames as input.

Did you notice that this code used the double equal operator, `==`? `==` is one of R's logical comparison operators. Comparison operators are key to using `filter()`, so let's take a look at them.

## Logical Comparisons

### Comparison operators

R provides a suite of comparison operators that you can use to compare values: `>`, `>=`, `<`, `<=`, `!=` (not equal), and `==` (equal). Each creates a logical test. For example, is `pi` greater than three?

```{r}
pi > 3
```
